---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# brms.mmrm

<!--
[![CRAN](https://www.r-pkg.org/badges/version/brms.mmrm)](https://CRAN.R-project.org/package=brms.mmrm)
-->
[![status](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![check](https://github.com/RConsortium/brms.mmrm/workflows/check/badge.svg)](https://github.com/RConsortium/brms.mmrm/actions?query=workflow%3Acheck)
[![cover](https://github.com/RConsortium/brms.mmrm/workflows/cover/badge.svg)](https://github.com/RConsortium/brms.mmrm/actions?query=workflow%3Acover)
[![lint](https://github.com/RConsortium/brms.mmrm/workflows/lint/badge.svg)](https://github.com/RConsortium/brms.mmrm/actions?query=workflow%3Alint)

The [mixed model for repeated measures (MMRM)](https://link.springer.com/article/10.1177/009286150804200402) is a popular model for longitudinal clinical trial data with continuous endpoints, and [`brms`](https://paul-buerkner.github.io/brms/) is powerful and versatile package for fitting Bayesian regression models. The `brms.mmrm` R package leverages [`brms`](https://paul-buerkner.github.io/brms/) to run [MMRMs](https://link.springer.com/article/10.1177/009286150804200402), and it supports a simplified interfaced to reduce difficulty and align with best practices for the life sciences.

## Installation

Type | Source | Command
---|---|---
Release | CRAN | `install.packages("brms.mmrm")`
Development | GitHub | `remotes::install_github("RConsortium/brms.mmrm")`
Development | RConsortium | `install.packages("brms.mmrm", repos = "https://RConsortium.r-universe.dev")`

## Documentation

The documentation website at <https://RConsortium.github.io/brms.mmrm> has a complete function reference and tutorial vignettes.

## Disclaimer

This package is still in its early development phase, which means:

1. It has not been validated for serious analyses, and
2. The interface and output may still change in ways that are not compatible with previous versions.

## Help

Please report questions and problems as [GitHub discussions](https://github.com/RConsortium/brms.mmrm) and [GitHub issues](https://github.com/RConsortium/brms.mmrm), respectively.

## Thanks

Thanks to the [ASA Biopharmaceutical Section Software Engineering Working Group](https://rconsortium.github.io/asa-biop-swe-wg/) and [R Consortium](https://www.r-consortium.org/) for providing professional networks to recruit skilled statisticians and developers.

## Code of conduct

Please note that the brms.mmrm project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Citation

```{r, warning = FALSE, comment = NA_character_, echo = FALSE}
citation("brms.mmrm")
```

## References

* Paul-Christian Bürkner (2017). brms: An R Package for Bayesian
  Multilevel Models Using Stan.
  Journal of Statistical Software, 80(1), 1-28.
  doi:10.18637/jss.v080.i01
* Mallinckrodt, C.H., Lane, P.W., Schnell, D. et al.
  Recommendations for the Primary Analysis of Continuous Endpoints
  in Longitudinal Clinical Trials.
  Ther Innov Regul Sci 42, 303–319 (2008).
  doi:10.1177/009286150804200402
