---
title: "Meta-analytic predictive priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meta-analytic predictive priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(brms)
library(brms.mmrm)
library(dplyr)
library(mmrm)
library(RBesT)
set.seed(0L)
```

A meta-analytic predictive prior is a type of historical borrowing approach that uses data from one or more previous studies to build a prior distribution for parameters of interest in a new study (@spiegelhalter2004, @neuenschwander2010, @schmidli2014). 
This vignette demonstrates how to supply a multivariate robust meta-analytic predictive prior to a Bayesian MMRM fitted with `brms.mmrm`.

## Data of the current study

We use the FEV1 data from the `mmrm` package, which contains measurements of forced expiratory volume in 1 second (FEV1) for patients with chronic obstructive pulmonary disease (COPD) at multiple visits.

```{r}
library(brms.mmrm)
library(dplyr)
data(fev_data, package = "mmrm")
data_current <- fev_data |>
  mutate(FEV1_CHG = FEV1 - FEV1_BL) |>
  brm_data(
    outcome = "FEV1_CHG",
    group = "ARMCD",
    time = "AVISIT",
    patient = "USUBJID",
    reference_group = "PBO"
  ) |>
  brm_data_chronologize(order = "VISITN") |>
  brm_archetype_effects(intercept = FALSE)
data_current
```

We are using a treatment effect informative prior archetype that invites the user to specify an informative prior on the placebo group mean at each study visit.
For more details on informative prior archetypes, see `vignette("archetypes", package = "brms.mmrm")`.

```{r}
summary(data_current)
```

## Benchmark analysis with a diffuse prior

As a basis for comparison, we first fit a Bayesian MMRM with a diffuse prior:

```{r, output = FALSE, message = FALSE, results = "hide"}
fit_diffuse <- brm_model(
  data = data_current,
  formula = brm_formula(data_current),
  refresh = 0L
)
```

The estimated marginal means closely match the analogous data summaries:

```{r}
draws_diffuse <- brm_marginal_draws(fit_diffuse)
summaries_diffuse <- brm_marginal_summaries(draws_diffuse)
summaries_data <- brm_marginal_data(data_current)
```

```{r}
brm_plot_compare(data = summaries_data, fit_diffuse = summaries_diffuse)
```

Suppose the trial is designed to declare efficacy if the posterior probability of observing a treatment effect at visit 4 ($\delta$) of at least 4 units of FEV1 is at least 0.8:

$$
\begin{aligned}
P(\delta > 4 \ \mid \ \text{data}) > 0.85
\end{aligned}
$$

The posterior probability undershoots the efficacy threshold, which is unsurprising because the dataset has few patients and MMRMs with diffuse priors are weak.

```{r}
draws_diffuse |>
  brm_marginal_probabilities(direction = "greater", threshold = 4) |>
  filter(time == "VIS4") |>
  pull(value)
```

## Constructing the robust MAP prior

Suppose we have a wealth of historical summary-level placebo data on FEV1 at visit 4 in similar studies:

```{r}
data_historical_visit4 <- tibble::tribble(
  ~study   , ~mean , ~sd   , ~patients ,
  "study1" , 8.16  , 10.01 ,       437 ,
  "study2" , 8.45  ,  9.87 ,       558 ,
  "study3" , 7.34  , 12.33 ,       489 ,
  "study4" , 6.87  , 14.44 ,       320 ,
  "study5" , 7.00  , 12.07 ,       491 ,
  "study6" , 7.10  , 11.11 ,       574
) |>
  mutate(se = sd / sqrt(patients))
```

We use `gMAP()` from `RBesT` to fit this summary data with a meta-analytic predictive model, then `automixfit()` to approximate the MAP posterior as a mixture of normals, and finally `robustify()` to add a weakly informative component that protects against prior-data conflict (@schmidli2014).
This robustified MAP posterior will serve as the MAP prior for the Bayesian MMRM downstream.^[See the documentation of `gMAP()`, `automixfit()` and `robustify()` in `RBesT` for details on how to do this in general.]

```{r, results = "hide", message = FALSE}
library(RBesT)
map_mcmc <- gMAP(
  cbind(mean, se) ~ 1 | study,
  data = data_historical_visit4,
  family = gaussian,
  tau.dist = "HalfNormal",
  tau.prior = 1,
  beta.prior = cbind(0, 100)
)

map_mix <- automixfit(map_mcmc)

map_prior <- robustify(
  map_mix,
  weight = 0.2,
  mean = 0,
  sigma = mean(data_historical_visit4$sd)
)
```

```{r}
map_prior
```

## Converting the prior for use in `brms.mmrm`

We assign robust MAP prior to the placebo group mean at visit 1:^[See the documentation of `RBesT::mixstanvar()`]

```{r}
prior <- brm_prior_label(
  code = "mixnorm(map_w, map_m, map_s)",
  group = "PBO",
  time = "VIS4"
) |>
  brm_prior_archetype(archetype = data_current)

prior
```

We then use `RBesT::mixstanvar()` to tell the Stan code in `brms` how to interpret `mixnorm()`.
Below, the name "map" has to align with the prefix "map" in the call to `mixnorm()` above.^[The suffixes "_w", "_m", and "_s" in `mixnorm()` are hard-coded.]

```{r}
stanvars <- mixstanvar(map = map_prior)
```

## Fitting the Bayesian MMRM with the MAP prior

We simply plug `prior` and `stanvars` into the call to `brms.mmrm::brm_model()`:

```{r, output = FALSE, message = FALSE, results = "hide"}
fit_map <- brm_model(
  data = data_current,
  formula = brm_formula(data_current),
  prior = prior,
  stanvars = stanvars,
  refresh = 0L
)
```

The model with the MAP prior has a more precise estimate of the placebo group mean at the final visit.

```{r}
draws_map <- brm_marginal_draws(fit_map)
summaries_map <- brm_marginal_summaries(draws_map)
```

```{r}
brm_plot_compare(
  data = summaries_data,
  fit_diffuse = summaries_diffuse,
  fit_map = summaries_map
)
```

With this added precision, we meet the efficacy threshold:

```{r}
draws_map |>
  brm_marginal_probabilities(direction = "greater", threshold = 4) |>
  filter(time == "VIS4") |>
  pull(value)
```

In other trials, the MAP prior may have the opposite effect.
To avoid human decision-making bias, it is important to pre-specify the analysis model used to evaluate the efficacy rule.

## Multivariate mixture priors

```{r, eval = FALSE}
prior <- brm_prior_label(
  code = "mixnorm(control_w, control_m, control_sigma_L)",
  group = "PBO"
)
```
