---
title: "Meta-analytic predictive priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meta-analytic predictive priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(brms)
library(brms.mmrm)
library(dplyr)
library(mmrm)
library(RBesT)
set.seed(0L)
```

A meta-analytic predictive prior is a type of historical borrowing approach that uses data from one or more previous studies to build a prior distribution for parameters of interest in a new study (@spiegelhalter2004, @neuenschwander2010, @schmidli2014). 
This vignette demonstrates how to supply a multivariate robust meta-analytic predictive prior to a Bayesian MMRM fitted with `brms.mmrm`.

## Data of the current study

We use the FEV1 data from the `mmrm` package, which contains measurements of forced expiratory volume in 1 second (FEV1) for patients with chronic obstructive pulmonary disease (COPD) at multiple visits.

```{r}
library(brms.mmrm)
library(dplyr)
data(fev_data, package = "mmrm")
data_current <- fev_data |>
  mutate(FEV1_CHG = FEV1 - FEV1_BL) |>
  brm_data(
    outcome = "FEV1_CHG",
    group = "ARMCD",
    time = "AVISIT",
    patient = "USUBJID",
    reference_group = "PBO"
  ) |>
  brm_data_chronologize(order = "VISITN") |>
  brm_archetype_effects(intercept = FALSE)
data_current
```

We are using a treatment effect informative prior archetype that invites the user to specify an informative prior on the placebo group mean at each study visit.
For more details on informative prior archetypes, see `vignette("archetypes", package = "brms.mmrm")`.

```{r}
summary(data_current)
```

## Benchmark analysis with a non-informative prior

Let's analyze this data with a non-informative prior first, to establish a baseline for comparison.

```{r, output = FALSE, message = FALSE, results = "hide"}
formula <- brm_formula(data_current)
fit_noninformative <- brm_model(
  data = data_current,
  formula = formula,
  refresh = 0L
)
```

```{r}
draws_noninformative <- brm_marginal_draws(fit_noninformative)
summaries_noninformative <- brm_marginal_summaries(draws_noninformative)
summaries_data_current <- brm_marginal_data(data_current)
brm_plot_compare(
  data_current = summaries_data_current,
  fit_noninformative = summaries_noninformative
)
```

Suppose the study was designed to declare efficacy if the posterior probability of observing a treatment effect at the final visit ($\delta$) greater than 4 units of FEV1 is greater than 75%:

$$
\begin{aligned}
P(\delta > 4 \ | \ \text{data}) > 0.75
\end{aligned}
$$

We see a posterior probability that barely meets this threshold:

```{r}
brm_marginal_probabilities(
  draws_noninformative,
  direction = "greater",
  threshold = 4
) |>
  filter(time == "VIS4") |>
  pull(value)
```

## Historical data

For demonstration purposes, we posit summary-level historical datasets on visit 4 that mimic the structure of the current study data but show a higher placebo response.^[In practice, you would use actual historical data from previous studies that are relevant to your current study.]

```{r}
data_historical_visit4 <- tibble::tribble(
  ~study   , ~mean , ~sd   , ~patients ,
  "study1" , 12.16 , 10.01 ,       137 ,
  "study2" , 13.45 ,  9.87 ,       158 ,
  "study3" , 11.34 , 12.33 ,        89 ,
  "study4" , 10.87 , 14.44 ,       120 ,
  "study5" , 12.00 , 12.07 ,        91 ,
  "study6" , 13.10 , 11.11 ,        74
) |>
  mutate(se = sd / sqrt(patients))
```

## Constructing the robust MAP prior

We use `gMAP()` from `RBesT` to fit a meta-analytic predictive model to the historical summary data, then `automixfit()` to approximate the MAP posterior as a mixture of normals, and finally `robustify()` to add a weakly informative component that protects against prior-data conflict (@schmidli2014).
This robustified MAP posterior will serve as the MAP prior for the Bayesian MMRM downstream.

```{r, results = "hide", message = FALSE}
library(RBesT)
map_mcmc <- gMAP(
  cbind(mean, se) ~ 1 | study,
  data = data_historical_visit4,
  family = gaussian,
  tau.dist = "HalfNormal",
  tau.prior = 1,
  beta.prior = cbind(0, 100)
)

map_mix <- automixfit(map_mcmc)

map_prior <- robustify(
  map_mix,
  weight = 0.2,
  mean = 0,
  sigma = mean(data_historical_visit4$sd)
)
```

```r
map_prior
```

## Analysis

